Functional timeseries reveal no systemic shifts for vertebrate communities

### 2021-04-21 paper/manuscript.Rmd
Writing group feedback

Ecological communities are experiencing unprecedented change as a result
of anthropogenic pressures such as climate change, land use change, and
invasive species. Impacts of these pressures are well documented at a
global scale by an accelerating global extinction rate [@barnosky2011],
and fundamental changes in some of the most well-studied systems (e.g.
coral bleaching). At the local scale however, species diversity tells a
different story. Recent synthesis of local trends in biodiversity over
time have found no net change in local biodiversity despite ongoing
species turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017]
and evidence of a decoupling of community composition from richness
[@brose2016; @gotelli2017] . Potential hypothesis for the balance of
etc..<!-- potential indication of biological homogenization due to shifting ranges and/or invasive species-->
While communities are clearly changing, our most common species-based
approaches do not fully capture the nature of that change.

Functional diversity offers a potentially powerful alternative for
detecting and describing community change by providing a mechanistic
link between the organisms in a system and the processes they perform
[@mcgill2006]. By summarizing functional traits rather than species,
functional diversity metrics capture the disproportionate impact of
losses or gains of functionally unique species. Functional diversity
metrics are therefore particularly well suited for addressing change in
communities with constant species richness.

Loss of functional diversity is also linked to impacts on critical
ecosystem and community processes. For example, there is increasing
evidence that functional diversity is a better predictor of ecosystem
function than species-based metrics [@cadotte2011; @gagic2015], and that
different facets of functional diversity play essential roles in
maintaining ecosystem stability [@craven2018; @morin2014]. Further,
maintenance of functional diversity is key for maximizing adaptive
potential of communities under climate change.

@@

Assessments of broad-scale temporal change in functional diversity have
previously been limited by a lack of functional trait data. The majority
of work has therefore focused largely on system-specific studies with
traits collected *in situ*. Ongoing efforts to assemble functional
traits for a variety of taxa have made synthesis of existing community
assemblage data and functional traits possible for the first time,
providing initial looks at how functional diversity changes on a
broad-scale for specific taxa (e.g. fish[@trindade-santos2020], birds
[@jarzyna2016; @barnagaud2017]).
<!-- so what do these studies say? does that go here? --> However, to
date there has been no cross-taxa assessment of temporal functional
change for a broad geographic and taxonomic extent.

Here we perform the first multi-taxa, multi-system assessment of
functional diversity change through time. We assess thousands of mammal,
bird, and amphibian functional diversity timeseries to 1) determine how
functional diversity change compares to species change and 2) assess the
types of functional shifts common across communities.
<!-- While our data represent the largest collection of functional diversity timeseries to date, they are not a representative sample of systems experiencing biodiversity change. We do not therefore seek to classify the global trend in functional diversity change, but rather assess the utility of a functional lens in understanding biodiversity change.-->

@@

We obtained mammal, bird, and amphibian timeseries from the Biotime
database, a global repository of high quality community assemblage
timeseries. All studies included in the database follow consistent
sampling protocols and represent full assemblages rather than population
of single species. [@dornelas2018]. Following best practices for the
database [@blowes2019] studies with multiple sample locations were split
into individual timeseries following a standardized scale. Scale was set
by a global grid with cell size determined based on the sample extent of
studies with only a single location (see [@dornelas2018] for details on
how sample extents were defined), with the area of each cell set to one
standard deviation away from the mean of the single extent locations.
All samples from a study within a single cell were considered to be a
single timeseries, and species abundances were combined for all samples.

We gathered trait data from the Elton Trait Database (mammals and birds;
[@wilman2014]) and Amphibio (amphibians, [@oliveira2017]). These
databases include species-level means for traits that best define
species' function in the community including body size, diet, and
behavioral characteristics <!-- what's in the amphibian database? -->.
For the full list of traits included in the analysis see the
supplemental materials.
<!-- limitations of species level trait data? -->

In order to ensure taxonomic consistency across datasets, Biotime
species were paired with trait data based on their species identifier
from the Integrated Taxonomic Information System database (citation),
obtained through the taxadb R package [@norman2020]. If more than one
species in the assemblage data resolved to the same identifier,
observations were considered the same species. For trait data, traits
for all species of the same identifier were averaged. Only studies with
at least 75% trait coverage were included. In order to have a sufficient
number of species to calculate functional diversity metrics years with
fewer than 4 species observed were excluded.

@@

Our final dataset included 2439 timeseries from 54 studies in 21
countries and 15 biomes. Only four studies (consisting of 11 timeseries)
come from Amphibian studies due to the limited availability of amphibian
timeseries and low species richness values for assemblages. For a full
breakdown of studies and their characteristics see the supplement.

@@

We calculated yearly metrics of functional and species diversity for
each timeseries. Species-based metrics include species richness (*S*)
and Jaccard similarity (*J*) as a measure of turnover. Jaccard
similarity was calculated relative to the first observed year for a
timeseries. A negative trend would therefore indicate increasing
turnover.

@@

We estimated overall trends for each diversity metric using a Linear
Mixed Effects model with a random slope and intercept for each
timeseries nested within study. Differences in trends by taxa, biome,
and realm were each modeled individually as covariates in models of the
same structure.

@@

We obtained study-level estimates of temporal change from the Best
Linear Unbiased Predictors (BLUP's) for each overall trend model and
assessed them in two ways. First, we looked at pairwise relationships
between all metrics to assess what kinds of community shifts are
commonly found together. We obtained correlations and bootstrapped
p-values using *corr.test* and *corCI* functions in the *psych* R
package (citation). Second, we assessed the relationship between all
metrics simultaneously by performing a cluster analysis which grouped
each study based on their similarities across all metrics. We used a
hierarchical clustering algorithm through the function *hclust* from the
R package *stats.* We included all standardized functional diversity
metrics, species richness, and Jaccard similarity in the analysis. We
report a qualitative interpretation of the groups from the cluster
analysis.

@@

At the study level, the majority of studies (78%) experienced a
significant temporal change in species richness and all functional
diversity metrics. 92% of sites experienced a significant trend for
species richness, 88% for functional functional richness, 96% for
functional divergence, 88% for functional evenness, and 35% for Jaccard
similarity. Only four studies experienced no significant trend in
species richness and were also characterized by significantly decreasing
functional evenness. There was only one significant pairwise correlation
between the metrics, a positive relationship between Functional
Divergence and Functional Evenness, indicating that studies experiencing
an increase in Divergence frequently experience an increase in Evenness.

Cluster analysis grouped all but a few studies into a cluster
characterized by non-significant or very small significant slopes for
all metrics. We identified four other clusters containing six studies
that were characterized by outlier trends in at least one of the metrics
and near-zero trends in the rest. The first contained two studies with
the most strongly decreasing evenness. The second, two studies with the
most strongly increasing functional divergence and increasing functional
richness. The third contained one study with strongly decreasing
functional divergence and richness, and the fourth strongly decreasing
species richness with increasing functional richness.

@@

Our study represents the largest broad-scale multi-taxa assessment of
functional change through time, giving a first look at overall trends in
functional diversity and the range of changes happening in mammal, bird,
and amphibian communities. Consistent with previous species-based
syntheses, we found no overall trend in species diversity accompanied by
increasing turnover through time [@dornelas2018]. While nearly all of
the studies included in our data experienced a significant increase or
decrease in species richness, they were observed in equal numbers
resulting in an overall trend not significantly different from zero.

@@

Individual studies in our dataset demonstrate the variety of species and
functional shifts that can happen concurrently in a community. With only
one significant pairwise relationship between Functional Divergence and
Functional Evenness, all other possible combinations of species and
functional change were observed in at least one study. Notably, there is
no relationship between species richness change and change in functional
richness, meaning that even for studies that may be experiencing species
losses there is not necessarily an accompanying loss of functional
richness and in some cases there may even be an increase. These
communities are gaining new species that are functional distinct enough
to make up for functional losses due to species losses, a pattern that
would be further magnified if the species lost are not functionally
distinct members of the assemblage.

@@

Our results have important implications for how we approach work linking
species loss to ecosystem processes, which often implicitly assume
concurrent functional loss. There is no evidence of directional
functional loss or even functional shifts accompany species loss in
these natural communities. Building on a rich body of work linking
species diversity changes to changes in ecosystem function, resilience,
and stability (citations), it will be essential to tease apart effects
due to species change from those due to functional shifts if we want to
infer impacts in natural communities.

>>>>

Ecological communities are experiencing unprecedented change as a result
of anthropogenic pressures such as climate change, land use change, and
invasive species. Impacts of these pressures are well documented at a
global scale by an accelerating global extinction rate [@barnosky2011],
and fundamental changes in some of the most well-studied systems (e.g.
coral bleaching). At the local scale however, species diversity tells a
different story. Recent syntheses of local trends in biodiversity over
time have found no net change in local species diversity despite ongoing
turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and
evidence of a significant shifts in community composition underlying
consistent species richness [@brose2016; @gotelli2017] .
<!-- potential indication of biological homogenization due to shifting ranges and/or invasive species-->
While communities are clearly changing, our most common species-based
approaches do not fully capture the nature of that change.

Functional diversity offers a potentially powerful alternative for
detecting and describing community change by providing a mechanistic
link between the organisms in a system and the processes they perform
[@mcgill2006]. By describing the functional trait space rather than
species, functional diversity metrics capture the disproportionate
impact of losses or gains of functionally unique species. Functional
diversity metrics are therefore particularly well-suited for addressing
change in communities with constant species richness.

Loss of functional diversity is also linked to impacts on critical
ecosystem and community processes. For example, there is increasing
evidence that functional diversity is a better predictor of ecosystem
function than species-based metrics [@cadotte2011; @gagic2015], and that
different facets of functional diversity play essential roles in
maintaining ecosystem stability [@craven2018; @morin2014]. Further,
maintenance of functional diversity is key for maximizing adaptive
potential of communities under climate change (citation).

@@

Assessments of broad-scale temporal change in functional diversity have
previously been limited by a lack of functional trait data. The majority
of work has therefore focused largely on system-specific studies with
traits collected *in situ*. Ongoing efforts to assemble functional
traits for a variety of taxa have made synthesis of existing community
assemblage data and functional traits possible for the first time,
providing initial insights into the ways functional diversity changes on
a broad scale for specific taxa (e.g. fish[@trindade-santos2020], birds
[@jarzyna2016; @barnagaud2017]).
<!-- so what do these studies say? does that go here? --> However, to
date there has been no cross-taxa assessment of temporal functional
change for a broad geographic and taxonomic extent.

Here we perform the first multi-taxa, multi-system assessment of
functional diversity change through time. We assess thousands of mammal,
bird, and amphibian functional diversity timeseries to 1) determine how
functional shifts compares to species change and 2) assess the types of
functional shifts that are most common across communities.
<!-- While our data represent the largest collection of functional diversity timeseries to date, they are not a representative sample of systems experiencing biodiversity change. We do not therefore seek to classify the global trend in functional diversity change, but rather assess the utility of a functional lens in understanding biodiversity change.-->

@@

We obtained mammal, bird, and amphibian timeseries from the Biotime
database, a global repository of high quality community assemblage
timeseries. All studies included in the database follow consistent
sampling protocols and represent full assemblages rather than population
of single species [@dornelas2018]. Following best practices for the
database [@blowes2019], studies with multiple sample locations were
split into individual timeseries following a standardized spatial scale.
Scale was set by a global grid with cell size determined based on the
sample extent of studies with only a single location (see
[@dornelas2018] for details on how sample extents were defined), with
the area of each cell set to one standard deviation away from the mean
of the single extent locations. All samples from a study within a single
cell were considered to be a single timeseries, and species abundances
were combined for all samples.

We gathered trait data from the Elton Trait Database (mammals and birds;
[@wilman2014]) and Amphibio (amphibians, [@oliveira2017]). These
databases include species-level means for traits that best define
species' function in the community including body size, diet, and
behavioral characteristics<!-- what's in the amphibian database? -->.
For the full list of traits included in the analysis see the
supplemental materials.
<!-- limitations of species level trait data? -->

In order to ensure taxonomic consistency across datasets, Biotime
species were paired with trait data based on their species identifier
from the Integrated Taxonomic Information System database (citation),
obtained through the taxadb R package [@norman2020]. If more than one
species in the assemblage data resolved to the same identifier,
observations were considered the same species. For trait data, traits
for all species of the same identifier were averaged. Only studies with
at least 75% trait coverage were included. In order to have a sufficient
number of species to calculate functional diversity metrics, years with
fewer than 4 species observed were excluded.

@@

Our final dataset included 2,439 timeseries from 54 studies in 21
countries and 15 biomes <!-- list traits -->. Only four studies
(consisting of 11 timeseries) came from Amphibian studies due to the
limited availability of amphibian timeseries and low species richness
values for assemblages. For a full breakdown of studies and their
characteristics, see the supplement.

@@

We calculated yearly metrics of functional and species diversity for
each timeseries. Species-based metrics include species richness (*S*)
and Jaccard similarity (*J*) as a measure of turnover. Jaccard
similarity was calculated relative to the first observed year for a
timeseries. A negative trend in *J* would therefore indicate increasing
turnover.

@@

We estimated overall trends for each diversity metric using a Linear
Mixed Effects model with a random slope and intercept for each
timeseries nested within the study. Differences in trends by taxa,
biome, and realm were each modeled individually as covariates in models
of the same structure.

@@

We obtained study-level estimates of temporal change from the Best
Linear Unbiased Predictors (BLUPs) for each overall trend model and
assessed them in two ways. First, we looked at pairwise relationships
between all metrics to assess what kinds of community shifts are
commonly found together. We obtained correlations and bootstrapped
p-values using *corr.test* and *corCI* functions in the *psych* R
package (citation). Second, we assessed the relationship between all
metrics simultaneously by performing a cluster analysis which grouped
each study based on their similarities across all metrics. We used a
hierarchical clustering algorithm through the function *hclust* from the
R package *stats.* We included all standardized functional diversity
metrics, species richness, and Jaccard similarity in the analysis. We
report a qualitative interpretation of the groups from the cluster
analysis.

@@

At the study level, the majority of studies (78%) experienced a
significant temporal change in species richness and all functional
diversity metrics. 92% of sites experienced a significant trend for
species richness, 88% for functional functional richness, 96% for
functional divergence, 88% for functional evenness, and 35% for Jaccard
similarity. Only 4 (cite studies) out of 54 studies experienced no
significant trend in species richness and were also characterized by
significantly decreasing functional evenness. There was only one
significant pairwise correlation between the metrics, a positive
relationship between Functional Divergence and Functional Evenness,
indicating that communities experiencing an increase in Divergence
frequently experience an increase in Evenness.

Cluster analysis grouped all but a few studies into a cluster
characterized by non-significant or very small significant slopes for
all metrics. We identified four other clusters containing six studies
(list studies) that were characterized by outlier trends in at least one
of the metrics and near-zero trends in the rest. The first contained two
studies with the most strongly decreasing evenness. The second, two
studies with the most strongly increasing functional divergence and
increasing functional richness. The third contained one study with
strongly decreasing functional divergence and richness, and the fourth
strongly decreasing species richness with increasing functional
richness.

@@

Our study represents the largest broad-scale multi-taxa assessment of
functional change through time, giving a first look at overall trends in
functional diversity and the range of changes happening in mammal, bird,
and amphibian communities. Consistent with previous species-based
syntheses, we found no overall trend in species diversity accompanied by
increasing turnover through time [@dornelas2018]. While nearly all of
the studies included in our data experienced a significant increase or
decrease in species richness, trends were observed in equal numbers
resulting in an overall trend not significantly different from zero.

@@

Individual studies in our dataset demonstrate the variety of species and
functional shifts that can happen concurrently in a community. With only
one significant pairwise relationship between Functional Divergence and
Functional Evenness, all other possible combinations of species and
functional change were observed in at least one study. Notably, there is
no relationship between species richness change and change in functional
richness, meaning that even for studies that may be experiencing species
losses there is not necessarily an accompanying loss of functional
richness and in some cases there may even be an increase. These
communities are gaining new species that are functionally distinct
enough to make up for functional losses due to species losses. This
pattern would be further magnified if the species lost are not
functionally distinct members of the assemblage.

@@

Our results have important implications for how we approach work linking
species loss to ecosystem processes, which often implicitly assume
concurrent functional loss. There is no evidence of directional
functional loss or even functional shifts accompany species loss in
these natural communities.
<!-- clarify that this is about relationship between concurrent patterns in a single community -->
Building on a rich body of work linking species diversity changes to
changes in ecosystem function, resilience, and stability (citations), it
will be essential to tease apart effects due to species change from
those due to functional shifts if we want to infer impacts in natural
communities.
